---
title: "GRNAD-SLAM NTR model estimation"
author: "Inés Rivero-García"
date: "31/03/2026"
output: 
  html_document:
    theme: cerulean
    number_sections: true
    toc: yes
    toc_float:
      collapsed: true
---

# SLAM-seq Test: estimate the proportion of new mRNA at different labelling pulse lenghts
```{r setup}
knitr::opts_knit$set(root.dir = "/Volumes/INTENSO/PhD/cardio_slamseq/")
library(ggplot2)

condition_colors <- c("#BBBBBB", "#fa8072", "#ff0800", "#420d09")
names(condition_colors) <- c("Null", "6hours", "12hours", "24hours")

condition_colors2 <- c("#BBBBBB", "#fa8072", "#ff0800", "#420d09")
names(condition_colors2) <- c("0", "6", "12", "24")
```


```{r load_data}
slamseq_test <- read.csv("data/GRAND-SLAM/cardio_slamseq_test.binomEstimated.tsv", 
                    header = TRUE, sep = "\t")
slamseq_test$Condition <- c(0, 6, 12, 24)

# Prep df for model fitting
df <- slamseq_test %>%
  dplyr::transmute(
    labelling_time = Condition,
    prop_new_mrna = p_new
  )
```
```{r plot_data}
ggplot(df, aes(labelling_time, prop_new_mrna)) +
  geom_point(size = 3) +
  labs(x = "Pulse duration", y = "Proportion labelled mRNA") +
  theme_bw() +
  scale_y_continuous(limits = c(0,1))
```
## Fitting an assymptotic model
$prop\ new\ mrna = A(1-e^{-k*time})$

The fraction of labeled mRNA increases asymptotically, reaching a plateau of ~72%. The half-time of labeling is ~4 hours (this means after 4h around 36% of the mRNAs should be labelled. 36% is half of the max. proportion of labeled mRNA).
```{r fit_assymptotic_model}
#' Asymptotic model. Estimate all parameters, then check they make sense 
#' R0 should be near 0 (0h labelling time should have prop_new_mrna = 0)
#' Asym should be close to max(p_new)
fit_asym <- nls(prop_new_mrna ~ SSasymp(labelling_time, Asym, R0, lrc), data = df)
summary(fit_asym)
print(paste0("Half-time of labelling: ",
             round(log(2) / exp(coef(fit_asym)["lrc"])),
      " hours"))
```


## Fitting a Michaelis-Menten model
$v = {Vmax[S]} \over {Km + [S]}$, where $S$ is the proportion of new mRNA.

This model estimates higher max. prop mRNA, but also sets the half-time of labelling ($Km$) to around 4 hours.

```{r fit_mm_model}
#' Michaelis-Menten model
#' Vmax is the highest detectable prop new mRNA
#' Km is the time needed to reach 50% of the max prop new mRNA. Value taken from assymptotic fit.

fit_mm <- nls(
  prop_new_mrna ~ (Vmax * labelling_time) / (Km + labelling_time),
  data = df,
  start = list(Vmax = max(slamseq_test$p_new_upper), 
               Km = 4)
)
summary(fit_mm)
```


## Compare models
The Michaelis-Menten model has better AIC, although both models give a very similar prediction. 

```{r compare_models}
print(paste0("Assymptotic model AIC: ", AIC(fit_asym)))
print(paste0("Michaelis-Menten model AIC: ", AIC(fit_mm)))

rmse <- function(fit) {
  sqrt(mean(resid(fit)^2))
}

print(paste0("Assymptotic model RSEM: ", rmse(fit_asym)))
print(paste0("Michaelis-Menten model RSEM: ", rmse(fit_mm)))
```
```{r inspect_residuals}
df$res_asym <- resid(fit_asym)
df$res_mm   <- resid(fit_mm)

ggplot(df) +
  geom_point(aes(x = labelling_time, y = res_asym), color = "orange") +
  geom_point(aes(x = labelling_time, y = res_mm), color = "blue") +
  theme_bw()
```
```{r visual_fit}
newdat <- data.frame(labelling_time = seq(0, 24, length.out = 200))
newdat$pred_asym <- predict(fit_asym, newdata = newdat)
newdat$pred_mm   <- predict(fit_mm,   newdata = newdat)

models_plt <- ggplot() +
  geom_hline(color="#420d09", yintercept=max(df$prop_new_mrna), size = 0.15, linetype = 1) +
  geom_smooth(data = df,
              aes(x = labelling_time, y = prop_new_mrna, color = "LOESS"),
              method = "loess", 
              se=FALSE, 
              linetype = 2, 
              size = 0.5) +
  geom_line(data = newdat, 
            aes(x = labelling_time, y = pred_asym, color = "Asymptotic"),
            linetype = 3, 
            size = 0.5) +
  geom_line(data = newdat, 
            aes(x = labelling_time, y = pred_mm, color = "Michaelis-Menten"), 
            linetype = 4, 
            size = 0.5) +
  geom_point(data = df,
             aes(x = labelling_time,
                 y = prop_new_mrna),
             size = 3) +
  theme_bw() +
  ylim(c(0, 1)) +
  scale_x_continuous(limits=c(0,24), breaks = seq(0, 24, 4)) +
  scale_y_continuous(limits= c(0,1), breaks = seq(0, 1, 0.1))+
  scale_color_manual(name = "Model",
                     values = c("LOESS" = "black",
                                "Asymptotic" = "orange",
                                "Michaelis-Menten" = "darkorchid")) +
  xlab("4-thiouridine pulse length (h)") + ylab("Proportion of new mRNA")

pdf("figures/01_preprocessing/GRAND-SLAM/model_fitting_SlamSeqTest.pdf",
    height = 3, width = 5)
models_plt
dev.off()

models_plt
```
```{r assymptotic_plot}
assymptotic_plt <- ggplot() +
  geom_hline(color="#420d09", yintercept=max(df$prop_new_mrna), size = 0.15, linetype = 1) +
  geom_line(data = newdat, 
            aes(x = labelling_time, y = pred_asym, color = "Asymptotic"),
            linetype = 1, 
            size = 0.5) +
  geom_point(data = df,
             aes(x = labelling_time,
                 y = prop_new_mrna),
             size = 3) +
  theme_bw() +
  ylim(c(0, 1)) +
  scale_x_continuous(limits=c(0,24), breaks = seq(0, 24, 4)) +
  scale_y_continuous(limits= c(0,1), breaks = seq(0, 1, 0.1))+
  scale_color_manual(name = "Model",
                     values = c("Asymptotic" = "orange")) +
  xlab("4-thiouridine pulse length (h)") + ylab("Proportion of new mRNA")

pdf("figures/01_preprocessing/GRAND-SLAM/assymptotic_model_fitting_SlamSeqTest.pdf",
    height = 3, width = 5)
assymptotic_plt
dev.off()

assymptotic_plt
```
```{r mm_plot}
mm_plt <- ggplot() +
  geom_hline(color="#420d09", yintercept=max(df$prop_new_mrna), size = 0.15, linetype = 1) +
  geom_line(data = newdat, 
            aes(x = labelling_time, y = pred_mm, color = "Michaelis-Menten"),
            linetype = 1, 
            size = 0.5) +
  geom_point(data = df,
             aes(x = labelling_time,
                 y = prop_new_mrna),
             size = 3) +
  theme_bw() +
  ylim(c(0, 1)) +
  scale_x_continuous(limits=c(0,24), breaks = seq(0, 24, 4)) +
  scale_y_continuous(limits= c(0,1), breaks = seq(0, 1, 0.1))+
  scale_color_manual(name = "Model",
                     values = c("Michaelis-Menten" = "darkorchid")) +
  xlab("4-thiouridine pulse length (h)") + ylab("Proportion of new mRNA")

pdf("figures/01_preprocessing/GRAND-SLAM/michaelis-menten_model_fitting_SlamSeqTest.pdf",
    height = 3, width = 5)
mm_plt
dev.off()

mm_plt
```


# Relationship between mismatches
Checking that mismatches only appear in the expected AG positions.
```{r mismatch1, echo=FALSE}
mismatch_df <- read.csv("data/GRAND-SLAM/cardio_slamseq_test.mismatches.tsv",
                        header = TRUE, sep = "\t")

mismatch_df <- mismatch_df %>%
  filter(Orientation == "First") %>%
  filter(Category == "Exonic") %>%
  mutate(Proportion = Mismatches/Coverage,
         Mismatch = paste0(Genomic, ">", Read),
         Condition = factor(Condition, levels = c("Null", "6hours", "12hours", "24hours")))

mismatch_plt <- ggplot(mismatch_df, aes(x = Mismatch, y = Proportion)) + 
  geom_bar(stat = "identity") + 
  facet_wrap(~ Condition) +
  theme_bw() +
  labs(y = "Proportion of mismatches", x = " Mismatch (Genomic > Read)") +
  scale_y_continuous(limits = c(0, 0.1), breaks = seq(0, 0.1, 0.02))

pdf("figures/01_preprocessing/GRAND-SLAM/mismatches_proportion_SlamSeqTest.pdf",
    height = 4, width = 8)
mismatch_plt
dev.off()

mismatch_plt
```