---
title: 'SLAM-seq differential expression analysis: MYC vs GFP'
author: "Inés Rivero-García"
date: "08/04/2026"
output: 
  html_document:
      theme: cerulean
      toc: true
      toc_float: true
      number_sections: true
---

```{r setup}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = "/Volumes/INTENSO/PhD/cardio_slamseq/")

library(RColorBrewer)
library(dplyr)
library(DESeq2)
library(ggplot2)
library(circlize)
library(ComplexHeatmap)
library(ggrepel)
library(UpSetR)
library(fgsea)

# Color palette for samples
colors_conditions <- c("#117733", "#AA3377", "#EE6677", "#DDAA33")
names(colors_conditions) <- c("GFP", "YAP5SA", "MYC", "caERBB2")

# Color palette for inter-sample distance heatmap
colors_distances <- colorRampPalette( rev(brewer.pal(9, "Blues")) )(255)

# Define directories
results_dir <- "results/02_degs/MYC_vs_GFP/"
figs_dir <- "figures/02_degs/MYC_vs_GFP/"

```
```{r load_clean_data}
data <- read.delim("data/GRAND-SLAM/cardio_slamseq_experiment.tsv")

# Clean data to make it a little smaller and separate between total and labeled counts
total <- data %>%
  dplyr::select(Gene, Symbol, Length, MYC_1.Readcount, MYC_2.Readcount, MYC_3.Readcount, GFP_2.1.Readcount, GFP_2.2.Readcount, GFP_2.3.Readcount)

labeled <- total
NTR <- data %>%
  dplyr::select(MYC_1.MAP, MYC_2.MAP, MYC_3.MAP, GFP_2.1.MAP, GFP_2.2.MAP, GFP_2.3.MAP)
  
# Labeled counts = raw counts * median new-to-old ratio
labeled[, 4:ncol(labeled)] <- round(labeled[, 4:ncol(labeled)] * NTR)

# Remove rows with unmeasured NTRs
labeled <- labeled[complete.cases(labeled), ]
#labeled[is.na(labeled)] <- 0 # Interpret NA's NTR as 0 NTR.

# Collapse to gene symbol for interpretability
labeled <- labeled %>%
  group_by(Symbol) %>%
  summarise(
    Gene = dplyr::first(Gene),
    across(where(is.numeric), sum),
    .groups = "drop"
  )
labeled <- as.data.frame(labeled)
rownames(labeled) <- labeled$Symbol
labeled <- labeled %>% filter(Symbol != "", !is.na(Symbol))

total <- total %>%
  group_by(Symbol) %>%
  summarise(
    Gene = dplyr::first(Gene),                       
    across(where(is.numeric), sum),           
    .groups = "drop"
  )
total <- as.data.frame(total)
rownames(total) <- total$Symbol
total <- total %>% filter(Symbol != "", !is.na(Symbol))

# Make shorter colnames
colnames(total) <- sub(".Readcount", "", colnames(total))
colnames(labeled) <- sub(".Readcount", "", colnames(labeled))

# Create DF with sample metadata
coldata <- data.frame(Sample = colnames(total)[4:ncol(total)],
                      Condition = sub("\\_.*", "", colnames(total)[4:ncol(total)]),
                      Batch = c(rep(2, 3), rep(2, 3)))
coldata <- coldata %>%
  mutate(Condition = case_when(Condition == "MYC" ~ "MYC", TRUE ~ Condition)) %>%
  mutate(Condition = factor(Condition, levels = c("MYC", "GFP"))) %>%
  mutate(Batch = factor(Batch))
rownames(coldata) <- coldata$Sample

# Check that row and colnames match
all(rownames(coldata) == colnames(total)[4:ncol(total)])
all(rownames(coldata) == colnames(labeled)[4:ncol(labeled)])

coldata
```


# PCA and sample clustering
```{r prepare_DDSobjects}
# Prepare gene info
geneInfo_total <- data.frame(Gene = total$Gene,
                      Symbol = total$Symbol,
                      Length = total$Length)
rownames(geneInfo_total) <- geneInfo_total$Gene

geneInfo_labeled <- data.frame(Gene = labeled$Gene,
                      Symbol = labeled$Symbol,
                      Length = labeled$Length)
rownames(geneInfo_labeled) <- geneInfo_labeled$Gene

# Prepare count matrices
total <- total %>%
  dplyr::select(MYC_1, MYC_2, MYC_3, GFP_2.1, GFP_2.2, GFP_2.3)
labeled <- labeled %>%
  dplyr::select(MYC_1, MYC_2, MYC_3, GFP_2.1, GFP_2.2, GFP_2.3)

# Create DESeq objects with all samples.
dds.total <- DESeqDataSetFromMatrix(countData = round(total),
                                    colData = coldata,
                                    design = ~ Condition) 
dds.labeled <- DESeqDataSetFromMatrix(countData = round(labeled),
                                      colData = coldata, 
                                      design = ~ Condition)

# Adding addtional gene info.
mcols(dds.total) <- DataFrame(mcols(dds.total), geneInfo_total)
mcols(dds.labeled) <- DataFrame(mcols(dds.labeled), geneInfo_labeled)

# Pre-filtering to speed things up. At least 3 samples with more than 10 counts.
keep <- rowSums(counts(dds.total) >= 10) >= 3
dds.total <- dds.total[keep,]
keep <- rowSums(counts(dds.labeled) >= 10) >= 3
dds.labeled <- dds.labeled[keep,]

# VST normalization
vsd.total <- vst(dds.total, blind = FALSE)
vsd.labeled <- vst(dds.labeled, blind = FALSE)
```

```{r pca}
pca.total <- plotPCA(vsd.total, intgroup = "Condition", returnData = TRUE)
pca.labeled <- plotPCA(vsd.labeled, intgroup = "Condition", returnData = TRUE)
pca.total <- merge(pca.total, coldata, 
                         by.x = c("Condition", "name"), by.y = c("Condition", "Sample"))
pca.labeled <- merge(pca.labeled, coldata, 
                           by.x = c("Condition", "name"), by.y = c("Condition", "Sample"))

pca.total <- ggplot(pca.total, aes(x = PC1, y = PC2, color = Condition)) +
  geom_point(size = 3) + 
  theme_bw() +
  scale_color_manual(values = colors_conditions) +
  ggtitle("PCA with total reads") +
  theme(plot.title = element_text(hjust = 0.5, face = "bold"))

pca.labeled <- ggplot(pca.labeled, aes(x = PC1, y = PC2, color = Condition)) +
  geom_point(size = 3) + 
  theme_bw() +
  scale_color_manual(values = colors_conditions) +
  ggtitle("PCA with labeled reads") +
  theme(plot.title = element_text(hjust = 0.5, face = "bold"))

pdf(paste0(figs_dir, "MYC_vs_GFP_pca_total.pdf"), height = 2.5, width = 3)
pca.total
dev.off()

pdf(paste0(figs_dir, "MYC_vs_GFP_pca_labeled.pdf"), height = 2.5, width = 3)
pca.labeled
dev.off()

pca.total
pca.labeled
```

```{r sample_clustering}
sampleDists.vsd.total <- dist(t(assay(vsd.total)), method = "euclidean")
sampleDists.vsd.labeled <- dist(t(assay(vsd.labeled)), method = "euclidean")

sampleDistMatrix.vsd.total <- as.matrix(sampleDists.vsd.total)
sampleDistMatrix.vsd.labeled <- as.matrix(sampleDists.vsd.labeled)

# Column annotation
ha <- HeatmapAnnotation(Condition = coldata$Condition,
                        col = list(Condition = colors_conditions)
)

heatmap.total <- Heatmap(sampleDistMatrix.vsd.total,
        name = "Euclidean\ndistance",
        top_annotation = ha,
        column_title = "VST Total counts",
        col = colorRamp2(seq(min(sampleDistMatrix.vsd.total), max(sampleDistMatrix.vsd.total), length.out = length(colors_distances)), colors_distances),
        cluster_rows = as.dendrogram(hclust(sampleDists.vsd.total)),
        cluster_columns =  as.dendrogram(hclust(sampleDists.vsd.total)))

heatmap.labeled <- Heatmap(sampleDistMatrix.vsd.labeled,
        name = "Euclidean\ndistance",
        top_annotation = ha,
        column_title = "VST Labeled counts",
        col = colorRamp2(seq(min(sampleDistMatrix.vsd.labeled), max(sampleDistMatrix.vsd.labeled), length.out = length(colors_distances)), colors_distances),
        cluster_rows = as.dendrogram(hclust(sampleDists.vsd.labeled)),
        cluster_columns =  as.dendrogram(hclust(sampleDists.vsd.labeled)))


pdf(paste0(figs_dir, "MYC_vs_GFP_sample-distance_hmap_total.pdf"), height = 3.5, width = 4.5)
heatmap.total
dev.off()

pdf(paste0(figs_dir, "MYC_vs_GFP_sample-distance_hmap_labeled.pdf"), height = 3.5, width = 4.5)
heatmap.labeled
dev.off()

heatmap.total
heatmap.labeled
```


# Differential expression analysis
More genes are DE when we use the total RNA-seq vs the labeled RNA-seq. This is expected because the labeled RNA-seq only considers the RNA-seq synthesized during 4h.
```{r deseq2}
# Set GFP as reference
dds.total$Condition <- relevel(dds.total$Condition, ref = "GFP")
dds.labeled$Condition <- relevel(dds.labeled$Condition, ref = "GFP")

dds.total <- DESeq(dds.total)
dds.labeled <- DESeq(dds.labeled)

res.total <- results(dds.total, contrast = c("Condition", "MYC", "GFP"), alpha = 0.1)
res.labeled <- results(dds.labeled, contrast = c("Condition", "MYC", "GFP"), alpha = 0.1)

summary(res.total)
summary(res.labeled)

# Log2FC shrinkage for plotting and ranking
resLFC.total <- lfcShrink(dds.total, coef="Condition_MYC_vs_GFP", type="apeglm")
resLFC.labeled<- lfcShrink(dds.labeled, coef="Condition_MYC_vs_GFP", type="apeglm")

# MA plots
maplt.total <- plotMA(resLFC.total, ylim=c(-2,2), main = "MYC vs GFP, Total reads") # shrinking log2FC
maplt.labeled <- plotMA(resLFC.labeled, ylim=c(-2,2), main = "MYC vs GFP, Labeled reads") # shrinking log2FC

pdf(paste0(figs_dir, "MYC_vs_GFP_MA-plot_total.pdf"), height = 3, width = 3.5)
plotMA(resLFC.total, ylim=c(-2,2), main = "MYC vs GFP, Total reads") # shrinking log2FC
dev.off()

pdf(paste0(figs_dir, "MYC_vs_GFP_MA-plot_labeled.pdf"), height = 3, width = 3.5)
plotMA(resLFC.labeled, ylim=c(-2,2), main = "MYC vs GFP, Labeled reads") # shrinking log2FC
dev.off()

maplt.total
maplt.labeled
```


## Sanity checks
The VST normalization looks good.
```{r mean_sd_plot}
pdf(paste0(figs_dir, "MYC_vs_GFP_mean-sd-plot_total.pdf"), height = 3, width = 3.5)
vsn::meanSdPlot(assay(vsd.total))
dev.off()

pdf(paste0(figs_dir, "MYC_vs_GFP_mean-sd-plot_labeled.pdf"), height = 3, width = 3.5)
vsn::meanSdPlot(assay(vsd.labeled))
dev.off()

vsn::meanSdPlot(assay(vsd.total))
vsn::meanSdPlot(assay(vsd.labeled))
```

```{r cook_distance}
pdf(paste0(figs_dir, "MYC_vs_GFP_cook-distances-plot_total.pdf"), height = 3, width = 3.5)
boxplot(log10(assays(dds.total)[["cooks"]]), range=0, las=2, 
        main = "Cook distances, Total RNA")
dev.off()

pdf(paste0(figs_dir, "MYC_vs_GFP_cook-distances-plot_labeled.pdf"), height = 3, width = 3.5)
boxplot(log10(assays(dds.labeled)[["cooks"]]), range=0, las=2, 
        main = "Cook distances, Labeled RNA")
dev.off()

boxplot(log10(assays(dds.total)[["cooks"]]), range=0, las=2, 
        main = "Cook distances, Total RNA")

boxplot(log10(assays(dds.labeled)[["cooks"]]), range=0, las=2, 
        main = "Cook distances, Labeled RNA")
```


```{r dispersionestimates}
# per-gene dispersion estimates together with the fitted mean-dispersion relationship.
# black points that are circled in blue = outliers

pdf(paste0(figs_dir, "MYC_vs_GFP_dispersion-estimates-plot_total.pdf"), height = 3.5, width = 4)
plotDispEsts(dds.total, main = "Dispersion estimates Total RNA")
dev.off()

pdf(paste0(figs_dir, "MYC_vs_GFP_dispersion-estimates-plot_labeled.pdf"), height = 3.5, width = 4)
plotDispEsts(dds.labeled, main = "Dispersion estimates Labeled RNA")
dev.off()

plotDispEsts(dds.total, main = "Dispersion estimates Total RNA")
plotDispEsts(dds.labeled, main = "Dispersion estimates Labeled RNA")
```


As a sanity check, the histograms of p-value distributions have nice shapes.
```{r pvalhist}
pdf(paste0(figs_dir, "MYC_vs_GFP_pval-hist_total.pdf"), height = 3.5, width = 3.5)
hist(res.total[res.total$baseMean > 1, "pvalue"], main = "P-value histogram, total RNA", xlab = "p-value")
dev.off()

pdf(paste0(figs_dir, "MYC_vs_GFP_pval-hist_labeled.pdf"), height = 3.5, width = 3.5)
hist(res.labeled[res.labeled$baseMean > 1, "pvalue"], main = "P-value histogram, labeled RNA", xlab = "p-value")
dev.off()

hist(res.total[res.total$baseMean > 1, "pvalue"], main = "P-value histogram, total RNA", xlab = "p-value")
hist(res.labeled[res.labeled$baseMean > 1, "pvalue"], main = "P-value histogram, labeled RNA", xlab = "p-value")
```


```{r annotate, include=FALSE, message=FALSE, warning=FALSE}
res.total$Symbol <- rownames(res.total)
res.labeled$Symbol <- rownames(res.labeled)
resLFC.total$Symbol <- rownames(resLFC.total)
resLFC.labeled$Symbol <- rownames(resLFC.labeled)

res.total <- merge(as.data.frame(res.total), geneInfo_total, by.x = "Symbol", by.y = "Symbol", all.x = TRUE)
res.labeled <- merge(as.data.frame(res.labeled), geneInfo_labeled, by.x = "Symbol", by.y = "Symbol", all.x = TRUE)
resLFC.total <- merge(as.data.frame(resLFC.total), geneInfo_total, by.x = "Symbol", by.y = "Symbol", all.x = TRUE)
resLFC.labeled <- merge(as.data.frame(resLFC.labeled), geneInfo_labeled, by.x = "Symbol", by.y = "Symbol", all.x = TRUE) 


# Save results
write.table(res.total, paste0(results_dir, "MYC_vs_GFP_degs_total.tsv"),
            col.names = TRUE, row.names = FALSE, quote = FALSE, sep = "\t")
write.table(res.labeled, paste0(results_dir, "MYC_vs_GFP_degs_labeled.tsv"),
            col.names = TRUE, row.names = FALSE, quote = FALSE, sep = "\t")
write.table(resLFC.total, paste0(results_dir, "MYC_vs_GFP_degs_lfcShrinkage_total.tsv"),
            col.names = TRUE, row.names = FALSE, quote = FALSE, sep = "\t")
write.table(resLFC.labeled, paste0(results_dir, "MYC_vs_GFP_degs_lfcShrinkage_labeled.tsv"),
            col.names = TRUE, row.names = FALSE, quote = FALSE, sep = "\t")
saveRDS(dds.total, paste0(results_dir, "MYC_vs_GFP_dds-object_total.rds"))
saveRDS(dds.labeled, paste0(results_dir, "MYC_vs_GFP_dds-object_labeled.rds"))
```


## Visualizing differentially expressed genes
```{r volcano}
resLFC.total <- as.data.frame(resLFC.total)
resLFC.labeled <- as.data.frame(resLFC.labeled)

v1 <- ggplot(resLFC.total[complete.cases(resLFC.total),], 
             aes(x = log2FoldChange, y = -log10(padj))) +
  geom_point(aes(color = ifelse(padj < 0.1, "padj < 0.1", "padj >= 0.1"))) +
  geom_label_repel(aes(label=ifelse(Symbol == "Myc", "Myc", "")), max.overlaps = Inf, min.segment.length = 0.01)+
  theme_bw() +
  xlab("Log2FC") +
  ylab("-log10(adjusted p-value)") +
  ggtitle("Total reads") +
  scale_color_manual(name = "Adjusted p-value", values = c("#EE6677", "grey80")) +
  theme(plot.title = element_text(hjust = 0.5, face = "bold"), legend.position = "right")

v2 <- ggplot(resLFC.labeled[complete.cases(resLFC.labeled),], 
             aes(x = log2FoldChange, y = -log10(padj))) +
  geom_point(aes(color = ifelse(padj < 0.1, "padj < 0.1", "padj >= 0.1"))) +
  geom_label_repel(aes(label=ifelse(Symbol == "Myc", "Myc", "")), max.overlaps = Inf, min.segment.length = 0.01)+
  theme_bw() +
  xlab("Log2FC") +
  ylab("-log10(adjusted p-value)") +
  ggtitle("Labeled reads") +
  scale_color_manual(name = "Adjusted p-value", values = c("#EE6677", "grey80")) +
  theme(plot.title = element_text(hjust = 0.5, face = "bold"), legend.position = "right")

pdf(paste0(figs_dir, "MYC_vs_GFP_volcano_total.pdf"), height = 3, width = 4)
v1
dev.off()

pdf(paste0(figs_dir, "MYC_vs_GFP_volcano_labeled.pdf"), height = 3, width = 4)
v2
dev.off()

v1
v2
```

```{r GeneHeatmap}
DEG.total <- resLFC.total %>%
  filter(padj < 0.1) %>%
  pull(Symbol)
DEG.labeled <- resLFC.labeled %>%
  filter(padj < 0.1) %>%
  pull(Symbol)

keep.total <- which(rownames(vsd.total) %in% DEG.total)
keep.labeled <- which(rownames(vsd.labeled) %in% DEG.labeled)

scaled.total <- scale(t(assay(vsd.total)[keep.total,]))
scaled.labeled <- scale(t(assay(vsd.labeled)[keep.labeled,]))

annoDF <- as.data.frame(coldata$Condition)
colnames(annoDF) <- "Condition"
rownames(annoDF) <- rownames(coldata)
annoCol<-list(Condition=c(GFP="#117733", MYC="#EE6677"))

# Gene clusters
column_ha <- HeatmapAnnotation(Condition = annoDF$Condition, col = list(Condition = c(colors_conditions)))

total_deg_heatmap <- Heatmap(t(scaled.total), 
                             name = "z-score", 
                             show_row_names = FALSE, 
                             top_annotation = column_ha, 
                             show_column_names = FALSE)

labeled_deg_heatmap <- Heatmap(t(scaled.labeled), 
                             name = "z-score", 
                             show_row_names = FALSE, 
                             top_annotation = column_ha, 
                             show_column_names = FALSE)

pdf(paste0(figs_dir, "MYC_vs_GFP_degs_hmap_total.pdf"), height = 3, width = 3.5)
total_deg_heatmap
dev.off()

pdf(paste0(figs_dir, "MYC_vs_GFP_degs_hmap_labeled.pdf"), height = 3, width = 3.5)
labeled_deg_heatmap
dev.off()

total_deg_heatmap
labeled_deg_heatmap
```


## Comparison between SLAM-seq and total RNA
```{r comparison, echo=FALSE, warning=FALSE}
compareDF <- merge(resLFC.labeled, resLFC.total,
                   by = c("Gene", "Symbol", "Length"),
                   all = TRUE)
colnames(compareDF) <- c("Gene", "Symbol", "Length", "baseMean_labeled", 
                         "log2FoldChange_labeled", "lfcSE_labeled", 
                         "pvalue_labeled", "padj_labeled",
                         "baseMean_total", "log2FoldChange_total", "lfcSE_total",
                         "pvalue_total", "padj_total")
comparePlot <- ggplot(compareDF, aes(x = log2FoldChange_total, y = log2FoldChange_labeled)) +
  geom_point(color = "#EE6677") + 
  geom_abline(slope = 1) +
  scale_x_continuous(limits = c(-12, 12)) +
  scale_y_continuous(limits = c(-4, 4)) +
  theme_bw() +
  xlab("Log2FC with total reads") +
  ylab("Log2FC with labeled reads") +
  ggtitle("MYC vs GFP") +
  theme(plot.title = element_text(hjust = 0.5, face = "bold"))
  
pdf(paste0(figs_dir, "MYC_vs_GFP_degs_significant_corr_total_labeled.pdf"), height = 3, width = 3)
comparePlot
dev.off()

comparePlot
```


```{r compare_genedir}
compareDF$padj_labeled[is.na(compareDF$padj_labeled)] <- 1
compareDF$padj_total[is.na(compareDF$padj_total)] <- 1

# Prepare lists of gene categories
list_input <- list(totalUP = compareDF %>% filter(padj_total < 0.1 & log2FoldChange_total > 0) %>% pull(Symbol),
                   totalDOWN = compareDF %>% filter(padj_total < 0.1 & log2FoldChange_total < 0) %>% pull(Symbol),
                   #totalUPNS = compareDF %>% filter(padj_total > 0.1 & log2FoldChange_total > 0) %>% pull(Symbol),
                   #totalDOWNNS = compareDF %>% filter(padj_total > 0.1 & log2FoldChange_total < 0) %>% pull(Symbol),
                   labeledUP = compareDF %>% filter(padj_labeled < 0.1 & log2FoldChange_labeled > 0) %>% pull(Symbol),
                   labeledDOWN = compareDF %>% filter(padj_labeled < 0.1 & log2FoldChange_labeled < 0) %>% pull(Symbol)#,
                   #labeledUPNS = compareDF %>% filter(padj_labeled > 0.1 & log2FoldChange_labeled > 0) %>% pull(Symbol),
                   #labeledDOWNNS = compareDF %>% filter(padj_labeled > 0.1 & log2FoldChange_labeled < 0) %>% pull(Symbol)
                   )

upset_plt <- upset(fromList(list_input), 
                   order.by = "freq", 
                   main.bar.color = "#EE6677", 
                   nintersects = NA,
                   matrix.color = "#EE6677",
                   sets.bar.color = "#EE6677", 
                   show.numbers = TRUE,
                   text.scale = 1)

pdf(paste0(figs_dir, "MYC_vs_GFP_degs_significant_upset-plot_total_labeled.pdf"), height = 4, width = 3)
upset_plt
dev.off()

upset_plt
```
# Validation of targets
## Targets described in the Bywater et al, 2020 paper.
Paper: Bywater MJ, Burkhart DL, Straube J, Sabò A, Pendino V, Hudson JE, Quaife-Ryan GA, Porrello ER, Rae J, Parton JG, Kress TR, Amati B, Littlewood TD, Evan GI & Wilson CH. Reactivation of Myc transcription in the mouse heart unlocks its proliferative capacity. Nat Commun 2020, 11;1827. doi: [10.1038/s41467-020-15552-x](https://doi.org/10.1038/s41467-020-15552-x)
```{r Heatmap_validate_paper}
# Load DEGs genes in Bywater et al MYC overexpression paper
paperUP <- readRDS("data/External/Bywater2020_MYC_UpregulatedGenes.rds")
paperDOWN <- readRDS("data/External/Bywater2020_MYC_DownregulatedGenes.rds")

valid.total <- resLFC.total %>%
  filter(is.element(Symbol, c(paperUP, paperDOWN)))
valid.labeled <- resLFC.labeled %>%
  filter(is.element(Symbol, c(paperUP, paperDOWN)))

keep.total <- which(rownames(vsd.total) %in% valid.total$Symbol)
keep.labeled <- which(rownames(vsd.labeled) %in% valid.labeled$Symbol)

scaled.total <- scale(t(assay(vsd.total)[keep.total,]))
scaled.labeled <- scale(t(assay(vsd.labeled)[keep.labeled,]))

# Plot total hmap
genes <- colnames(scaled.total)
gene_status <- ifelse(genes %in% paperUP, "Upregulated", ifelse(genes %in% paperDOWN, "Downregulated", NA))
status_colors <- c("Upregulated" = "orange", "Downregulated" = "darkgreen")
row_ha_total <- rowAnnotation(Bywater2020 = gene_status, col = list(Bywater2020 = status_colors))

total_bywater_hmap <- Heatmap(t(scaled.total), 
        left_annotation = row_ha_total, 
        cluster_rows = TRUE,
        top_annotation = column_ha,
        show_row_names = FALSE,
        show_column_names = FALSE,
        name = "scaled counts")

# Plot labeled hmap
genes <- colnames(scaled.labeled)
gene_status <- ifelse(genes %in% paperUP, "Upregulated", ifelse(genes %in% paperDOWN, "Downregulated", NA))
status_colors <- c("Upregulated" = "orange", "Downregulated" = "darkgreen")
row_ha_labeled <- rowAnnotation(Bywater2020 = gene_status, col = list(Bywater2020 = status_colors))

labeled_bywater_hmap <- Heatmap(t(scaled.labeled), 
        left_annotation = row_ha_labeled, 
        cluster_rows = TRUE,
        top_annotation = column_ha,
        show_row_names = FALSE,
        show_column_names = FALSE,
        name = "scaled counts")

pdf(paste0(figs_dir, "MYC_vs_GFP_Bywater2020_hmap_total.pdf"), height = 4, width = 5)
total_bywater_hmap
dev.off()

pdf(paste0(figs_dir, "MYC_vs_GFP_Bywater2020_hmap_labeled.pdf"), height = 4, width = 5)
labeled_bywater_hmap
dev.off()

total_bywater_hmap
labeled_bywater_hmap
```
```{r volcano_validate_paper}
# Volcano with labels
volcano_bywater_total <- ggplot(resLFC.total[complete.cases(resLFC.total),], aes(x = log2FoldChange, y = -log10(padj))) +
  geom_point(aes(color = ifelse(padj < 0.1, "padj < 0.1", "padj >= 0.1"))) +
  geom_label_repel(aes(label=ifelse(Symbol %in% c(paperUP, paperDOWN), as.character(Symbol),'')),max.overlaps = 10)+
  theme_bw() +
  xlab("Log2FC") +
  ylab("-log10(adjusted p-value)") +
  ggtitle("Genes from Bywater et al 2020\nTotal RNA") +
  scale_color_manual(name = "Adjusted p-value", values = c("#EE6677", "grey80")) +
  theme(plot.title = element_text(hjust = 0.5, face = "bold"))

volcano_bywater_labeled <- ggplot(resLFC.labeled[complete.cases(resLFC.labeled),], aes(x = log2FoldChange, y = -log10(padj))) +
  geom_point(aes(color = ifelse(padj < 0.1, "padj < 0.1", "padj >= 0.1"))) +
  geom_label_repel(aes(label=ifelse(Symbol %in% c(paperUP, paperDOWN), as.character(Symbol),'')), max.overlaps = 20)+
  theme_bw() +
  xlab("Log2FC") +
  ylab("-log10(adjusted p-value)") +
  ggtitle("Genes from Bywater et al 2020\nLabeled RNA") +
  scale_color_manual(name = "Adjusted p-value", values = c("#EE6677", "grey80")) +
  theme(plot.title = element_text(hjust = 0.5, face = "bold"))

pdf(paste0(figs_dir, "MYC_vs_GFP_Bywater2020_volcano_total.pdf"), height = 3, width = 4)
volcano_bywater_total
dev.off()

pdf(paste0(figs_dir, "MYC_vs_GFP_Bywater2020_volcano_labeled.pdf"), height = 3, width = 4)
volcano_bywater_labeled
dev.off()

volcano_bywater_total
volcano_bywater_labeled
```


```{r ExprPlot_validate_paper}
paper_genes <- c(paperUP, paperDOWN)

# Select genes described in paper that are significant in the total or labeled experiment
signif_total <- resLFC.total %>% filter(padj < 0.1) %>% pull(Symbol)
signif_labeled <- resLFC.labeled %>% filter(padj < 0.1) %>% pull(Symbol)
signif_genes <- unique(c(signif_total, signif_labeled))
paper_genes <- intersect(paper_genes, signif_genes)

for(gene in paper_genes){
  if(gene %in% rownames(dds.total)){
    total_counts <- plotCounts(dds.total, gene = gene, intgroup = "Condition", returnData = TRUE)
    
    log2fc.total <- resLFC.total %>% filter(Symbol == gene) %>% pull(log2FoldChange) 
    padj.total <- resLFC.total %>% filter(Symbol == gene) %>% pull(padj)
    
    plot.total <- ggplot(total_counts, aes(x = Condition, y = count, color = Condition, fill = Condition)) +
      geom_boxplot(alpha = 0.5)+
      geom_point(position=position_jitter(w=0.1,h=0)) +
      scale_color_manual(values = colors_conditions) +
      scale_fill_manual(values = colors_conditions) +
      theme_bw() +
      xlab("Condition") + 
      ylab("Normalized counts") +
      ggtitle(paste0(gene, " (Total RNA)"))+
      labs(subtitle = paste0("Log2FC vs GFP = ", round(log2fc.total, 2), 
                             "; adj.p = ", round(padj.total, 3), ")"))+
      theme(plot.title = element_text(hjust = 0.5, face = "bold"), 
            plot.subtitle = element_text(hjust = 0.5),
            legend.position = "none")
  }
  
  if(gene %in% rownames(dds.labeled)){
    labeled_counts <- plotCounts(dds.labeled, gene = gene, intgroup = "Condition", returnData = TRUE)
    
    log2fc.labeled <- resLFC.labeled %>% filter(Symbol == gene) %>% pull(log2FoldChange) 
    padj.labeled <- resLFC.labeled %>% filter(Symbol == gene) %>% pull(padj)
    
    plot.labeled <- ggplot(labeled_counts, aes(x = Condition, y = count, color = Condition, fill = Condition)) +
      geom_boxplot(alpha = 0.5)+
      geom_point(position=position_jitter(w=0.1,h=0)) +
      scale_color_manual(values = colors_conditions) +
      scale_fill_manual(values = colors_conditions) +
      theme_bw() +
      xlab("Condition") + 
      ylab("Normalized counts") +
      ggtitle(paste0(gene, " (Labeled RNA)"))+
      labs(subtitle = paste0("Log2FC vs GFP = ", round(log2fc.labeled, 2), 
                             "; adj.p = ", round(padj.labeled, 3), ")"))+
      theme(plot.title = element_text(hjust = 0.5, face = "bold"), 
            plot.subtitle = element_text(hjust = 0.5),
            legend.position = "none")
  }
  
  pdf(paste0(figs_dir, "Bywater2020_genes_significant/MYC_vs_GFP_Bywater2020_", gene, "_boxplot.pdf"),
    height = 3, width = 7)
  print(cowplot::plot_grid(plotlist = list(plot.total, plot.labeled), ncol = 2))
  dev.off()
}
```


## Targets in the CollecTRI database
```{r Heatmap_validate_collectri}
collectri <- readRDS("data/External/CollecTRI_from_OmnipathR_Mouse_39849edges_20260407.rds")
collectriPOS <- collectri %>%
  filter(source_genesymbol == "Myc") %>%
  filter(consensus_stimulation == 1) %>%
  pull(target_genesymbol)
collectriNEG <- collectri %>%
  filter(source_genesymbol == "Myc") %>%
  filter(consensus_inhibition == 1) %>%
  pull(target_genesymbol)

valid.total <- resLFC.total %>%
  filter(is.element(Symbol, c(collectriPOS, collectriNEG)))
valid.labeled <- resLFC.labeled %>%
  filter(is.element(Symbol, c(collectriPOS, collectriNEG)))

keep.total <- which(rownames(vsd.total) %in% valid.total$Symbol)
keep.labeled <- which(rownames(vsd.labeled) %in% valid.labeled$Symbol)

scaled.total <- scale(t(assay(vsd.total)[keep.total,]))
scaled.labeled <- scale(t(assay(vsd.labeled)[keep.labeled,]))

# Plot total hmap
genes <- colnames(scaled.total)
gene_status <- ifelse(genes %in% collectriPOS, "Activated", ifelse(genes %in% collectriNEG, "Inhibited", NA))
status_colors <- c("Activated" = "orange", "Inhibited" = "darkgreen")
row_ha_total <- rowAnnotation(CollecTRI = gene_status, col = list(CollecTRI = status_colors))

total_collectri_hmap <- Heatmap(t(scaled.total), 
        left_annotation = row_ha_total, 
        cluster_rows = TRUE,
        top_annotation = column_ha,
        show_row_names = FALSE,
        show_column_names = FALSE,
        name = "scaled counts")

# Plot labeled hmap
genes <- colnames(scaled.labeled)
gene_status <- ifelse(genes %in% collectriPOS, "Activated", ifelse(genes %in% collectriNEG, "Inhibited", NA))
status_colors <- c("Activated" = "orange", "Inhibited" = "darkgreen")
row_ha_labeled <- rowAnnotation(CollecTRI = gene_status, col = list(CollecTRI = status_colors))

labeled_collectri_hmap <- Heatmap(t(scaled.labeled), 
        left_annotation = row_ha_labeled, 
        cluster_rows = TRUE,
        top_annotation = column_ha,
        show_row_names = FALSE,
        show_column_names = FALSE,
        name = "scaled counts")

pdf(paste0(figs_dir, "MYC_vs_GFP_collectri_hmap_total.pdf"), height = 4, width = 5)
total_collectri_hmap
dev.off()

pdf(paste0(figs_dir, "MYC_vs_GFP_collectri_hmap_labeled.pdf"), height = 4, width = 5)
labeled_collectri_hmap
dev.off()

total_collectri_hmap
labeled_collectri_hmap
```
```{r volcano_validate_collectri}
# Volcano with labels
volcano_collectri_total <- ggplot(resLFC.total[complete.cases(resLFC.total),], aes(x = log2FoldChange, y = -log10(padj))) +
  geom_point(aes(color = ifelse(padj < 0.1, "padj < 0.1", "padj >= 0.1"))) +
  geom_label_repel(aes(label=ifelse(Symbol %in% c(collectriPOS, collectriNEG), as.character(Symbol),'')),max.overlaps = 50)+
  theme_bw() +
  xlab("Log2FC") +
  ylab("-log10(adjusted p-value)") +
  ggtitle("Myc targets in CollecTRI\nTotal RNA") +
  scale_color_manual(name = "Adjusted p-value", values = c("#EE6677", "grey80")) +
  theme(plot.title = element_text(hjust = 0.5, face = "bold"))

volcano_collectri_labeled <- ggplot(resLFC.labeled[complete.cases(resLFC.labeled),], aes(x = log2FoldChange, y = -log10(padj))) +
  geom_point(aes(color = ifelse(padj < 0.1, "padj < 0.1", "padj >= 0.1"))) +
  geom_label_repel(aes(label=ifelse(Symbol %in% c(collectriPOS, collectriNEG), as.character(Symbol),'')), max.overlaps = 50)+
  theme_bw() +
  xlab("Log2FC") +
  ylab("-log10(adjusted p-value)") +
  ggtitle("Myc targets in CollecTRI\nLabeled RNA") +
  scale_color_manual(name = "Adjusted p-value", values = c("#EE6677", "grey80")) +
  theme(plot.title = element_text(hjust = 0.5, face = "bold"))

pdf(paste0(figs_dir, "MYC_vs_GFP_collectri_volcano_total.pdf"), height = 3, width = 4)
volcano_collectri_total
dev.off()

pdf(paste0(figs_dir, "MYC_vs_GFP_collectri_volcano_labeled.pdf"), height = 3, width = 4)
volcano_collectri_labeled
dev.off()

volcano_collectri_total
volcano_collectri_labeled
```
```{r ExprPlot_validate_collectri}
# Plot all detected genes because they are few

collectri_genes <- c(collectriPOS, collectriNEG)

for(gene in collectri_genes){
  if(gene %in% rownames(dds.total)){
    total_counts <- plotCounts(dds.total, gene = gene, intgroup = "Condition", returnData = TRUE)
    
    log2fc.total <- resLFC.total %>% filter(Symbol == gene) %>% pull(log2FoldChange) 
    padj.total <- resLFC.total %>% filter(Symbol == gene) %>% pull(padj)
    
    plot.total <- ggplot(total_counts, aes(x = Condition, y = count, color = Condition, fill = Condition)) +
      geom_boxplot(alpha = 0.5)+
      geom_point(position=position_jitter(w=0.1,h=0)) +
      scale_color_manual(values = colors_conditions) +
      scale_fill_manual(values = colors_conditions) +
      theme_bw() +
      xlab("Condition") + 
      ylab("Normalized counts") +
      ggtitle(paste0(gene, " (Total RNA)"))+
      labs(subtitle = paste0("Log2FC vs GFP = ", round(log2fc.total, 2), 
                             "; adj.p = ", round(padj.total, 3), ")"))+
      theme(plot.title = element_text(hjust = 0.5, face = "bold"), 
            plot.subtitle = element_text(hjust = 0.5),
            legend.position = "none")
  }
  
  if(gene %in% rownames(dds.labeled)){
    labeled_counts <- plotCounts(dds.labeled, gene = gene, intgroup = "Condition", returnData = TRUE)
    
    log2fc.labeled <- resLFC.labeled %>% filter(Symbol == gene) %>% pull(log2FoldChange) 
    padj.labeled <- resLFC.labeled %>% filter(Symbol == gene) %>% pull(padj)
    
    plot.labeled <- ggplot(labeled_counts, aes(x = Condition, y = count, color = Condition, fill = Condition)) +
      geom_boxplot(alpha = 0.5)+
      geom_point(position=position_jitter(w=0.1,h=0)) +
      scale_color_manual(values = colors_conditions) +
      scale_fill_manual(values = colors_conditions) +
      theme_bw() +
      xlab("Condition") + 
      ylab("Normalized counts") +
      ggtitle(paste0(gene, " (Labeled RNA)"))+
      labs(subtitle = paste0("Log2FC vs GFP = ", round(log2fc.labeled, 2), 
                             "; adj.p = ", round(padj.labeled, 3), ")"))+
      theme(plot.title = element_text(hjust = 0.5, face = "bold"), 
            plot.subtitle = element_text(hjust = 0.5),
            legend.position = "none")
  }
  
  pdf(paste0(figs_dir, "CollecTRI_genes_significant/MYC_vs_GFP_collectri_", gene, "_boxplot.pdf"),
    height = 3, width = 7)
  print(cowplot::plot_grid(plotlist = list(plot.total, plot.labeled), ncol = 2))
  dev.off()
}
```


# Enrichment analysis
```{r load_genesets}
reactome <- gmtPathways("data/External/MSigDB_Mmusculus/m2.cp.reactome.v2023.1.Mm.symbols.gmt")
go_bp <- gmtPathways("data/External/MSigDB_Mmusculus/m5.go.bp.v2023.1.Mm.symbols.gmt")
go_mf <- gmtPathways("data/External/MSigDB_Mmusculus/m5.go.mf.v2023.1.Mm.symbols.gmt")
go_cc <- gmtPathways("data/External/MSigDB_Mmusculus/m5.go.cc.v2023.1.Mm.symbols.gmt")
```
```{r prepare_ranks, include=FALSE}
ranks.total <- res.total$stat
names(ranks.total) <- res.total$Symbol
ranks.total <- sort(ranks.total)

ranks.labeled <- res.labeled$stat
names(ranks.labeled) <- res.labeled$Symbol
ranks.labeled <- sort(ranks.labeled)
```
```{r enrichment_reactome}
reactome.total <- fgsea(pathways = reactome, stats = ranks.total, minSize  = 10, maxSize  = 500)
reactome.labeled <- fgsea(pathways = reactome, stats = ranks.labeled, minSize  = 10, maxSize  = 500)

reactome.total$leadingEdge = sapply(reactome.total$leadingEdge, function(x) paste0(x, collapse = ";"))
reactome.labeled$leadingEdge = sapply(reactome.labeled$leadingEdge, function(x) paste0(x, collapse = ";"))

write.table(reactome.total, paste0(results_dir, "MYC_vs_GFP_gsea_reactome_total.tsv"), col.names = TRUE, row.names = FALSE, quote = FALSE, sep = "\t")
write.table(reactome.labeled, paste0(results_dir, "MYC_vs_GFP_gsea_reactome_labeled.tsv"), col.names = TRUE, row.names = FALSE, quote = FALSE, sep = "\t")

# Rank plot: label pathways with padj < 0.01 and |NES| > 2
reactome.total <- reactome.total %>%
  # Create signed log(padj) column for rank plot
  mutate(SignedLogPadj = -log10(padj)) %>%
  mutate(SignedLogPadj = ifelse(NES > 0, SignedLogPadj, -SignedLogPadj)) %>%
  # Create category column for coloring
  mutate(Category = case_when(
    padj < 0.01 & NES > 0 ~ "UP",
    padj < 0.01 & NES < 0 ~ "DOWN",
    TRUE ~ "ns")) %>%
  # Clean pathway names
  mutate(pathway_clean = pathway %>%
           stringr::str_remove("^REACTOME_") %>%
           stringr::str_replace_all("_", " ") %>%
           stringr::str_to_lower() %>%
           stringr::str_to_title())%>%
  # Only name pathways with padj < 0.01 and |NES| > 2
  mutate(pathway_plot = ifelse(abs(NES) > 2 & padj < 0.01, pathway_clean, ""))

reactome.labeled <- reactome.labeled %>%
  # Create signed log(padj) column for rank plot
  mutate(SignedLogPadj = -log10(padj)) %>%
  mutate(SignedLogPadj = ifelse(NES > 0, SignedLogPadj, -SignedLogPadj)) %>%
  # Create category column for coloring
  mutate(Category = case_when(
    padj < 0.01 & NES > 0 ~ "UP",
    padj < 0.01 & NES < 0 ~ "DOWN",
    TRUE ~ "ns")) %>%
  # Clean pathway names
  mutate(pathway_clean = pathway %>%
           stringr::str_remove("^REACTOME_") %>%
           stringr::str_replace_all("_", " ") %>%
           stringr::str_to_lower() %>%
           stringr::str_to_title()) %>%
  # Only name pathways with padj < 0.01 and |NES| > 2
  mutate(pathway_plot = ifelse(abs(NES) > 2 & padj < 0.01, pathway_clean, ""))

reactome_plt_total <- ggplot(reactome.total, aes(x = NES, y = SignedLogPadj, color = Category, label = pathway_plot)) +
  geom_point(size = 2) +
  geom_hline(yintercept = -log10(0.05), linetype='dotted', col = 'grey20') +
  geom_hline(yintercept = log10(0.05), linetype='dotted', col = 'grey20') +
  geom_label_repel(max.overlaps = 10) +
  theme_bw() +
  scale_color_manual(values = c("steelblue4", "grey70", "tomato3")) +
  xlab("Normalized Enrichment Score") +
  ylab("Signed -log10(adjusted p-value)") +
  ggtitle("Total RNA") +
  theme(plot.title = element_text(hjust = 0.5, face = "bold"))

reactome_plt_labeled <- ggplot(reactome.labeled, aes(x = NES, y = SignedLogPadj, color = Category, label = pathway_plot)) +
  geom_point(size = 2) +
  geom_hline(yintercept = -log10(0.05), linetype='dotted', col = 'grey20') +
  geom_hline(yintercept = log10(0.05), linetype='dotted', col = 'grey20') +
  geom_label_repel(max.overlaps = 10) +
  theme_bw() +
  scale_color_manual(values = c("steelblue4", "grey70", "tomato3")) +
  xlab("Normalized Enrichment Score") +
  ylab("Signed -log10(adjusted p-value)") +
  ggtitle("Labeled RNA") +
  theme(plot.title = element_text(hjust = 0.5, face = "bold"))

pdf(paste0(figs_dir, "MYC_vs_GFP_fgsea_reactome_ranks_total.pdf"), height = 6, width = 6)
reactome_plt_total
dev.off()

pdf(paste0(figs_dir, "MYC_vs_GFP_fgsea_reactome_ranks_labeled.pdf"), height = 6, width = 6)
reactome_plt_labeled
dev.off()

# Heatmap to compare labeled vs total
reactome.nes.total <- reactome.total %>%
  filter(padj < 0.01 & abs(NES) > 2) %>%
  select(pathway_clean, NES) %>%
  reshape2::melt()
reactome.nes.total$variable <- "Total"

reactome.nes.labeled <- reactome.labeled %>%
  filter(padj < 0.01 & abs(NES) > 2) %>%
  select(pathway_clean, NES) %>%
  reshape2::melt()
reactome.nes.labeled$variable <- "Labeled"

reactome.compare <- rbind(reactome.nes.total, reactome.nes.labeled)

reactome_compare_hmap <- ggplot(reactome.compare, aes(x = variable, y = pathway_clean, fill = value)) + 
  geom_tile(color = "black") + 
  theme_bw() + 
  scale_fill_gradient2(low = "steelblue4", mid = "white", high = "tomato3", name = "NES") +
  xlab("RNA") +
  ylab("REACTOME pathway")

pdf(paste0(figs_dir, "MYC_vs_GFP_fgsea_reactome_comparison_hmap.pdf"), height = 7, width = 10)
reactome_compare_hmap
dev.off()

reactome_plt_total
reactome_plt_labeled
reactome_compare_hmap
```


```{r enrichment_go_bp}
go_bp.total <- fgsea(pathways = go_bp, stats = ranks.total, minSize  = 10, maxSize  = 500)
go_bp.labeled <- fgsea(pathways = go_bp, stats = ranks.labeled, minSize  = 10, maxSize  = 500)

go_bp.total$leadingEdge = sapply(go_bp.total$leadingEdge, function(x) paste0(x, collapse = ";"))
go_bp.labeled$leadingEdge = sapply(go_bp.labeled$leadingEdge, function(x) paste0(x, collapse = ";"))

write.table(go_bp.total, paste0(results_dir, "MYC_vs_GFP_gsea_go-bp_total.tsv"), col.names = TRUE, row.names = FALSE, quote = FALSE, sep = "\t")
write.table(go_bp.labeled, paste0(results_dir, "MYC_vs_GFP_gsea_go-bp_labeled.tsv"), col.names = TRUE, row.names = FALSE, quote = FALSE, sep = "\t")

# Rank plot: label pathways with padj < 0.01 and |NES| > 2
go_bp.total <- go_bp.total %>%
  # Create signed log(padj) column for rank plot
  mutate(SignedLogPadj = -log10(padj)) %>%
  mutate(SignedLogPadj = ifelse(NES > 0, SignedLogPadj, -SignedLogPadj)) %>%
  # Create category column for coloring
  mutate(Category = case_when(
    padj < 0.01 & NES > 0 ~ "UP",
    padj < 0.01 & NES < 0 ~ "DOWN",
    TRUE ~ "ns")) %>%
  # Clean pathway names
  mutate(pathway_clean = pathway %>%
           stringr::str_remove("^GOBP_") %>%
           stringr::str_replace_all("_", " ") %>%
           stringr::str_to_lower() %>%
           stringr::str_to_title())%>%
  # Only name pathways with padj < 0.01 and |NES| > 2
  mutate(pathway_plot = ifelse(abs(NES) > 2 & padj < 0.01, pathway_clean, ""))

go_bp.labeled <- go_bp.labeled %>%
  # Create signed log(padj) column for rank plot
  mutate(SignedLogPadj = -log10(padj)) %>%
  mutate(SignedLogPadj = ifelse(NES > 0, SignedLogPadj, -SignedLogPadj)) %>%
  # Create category column for coloring
  mutate(Category = case_when(
    padj < 0.01 & NES > 0 ~ "UP",
    padj < 0.01 & NES < 0 ~ "DOWN",
    TRUE ~ "ns")) %>%
  # Clean pathway names
  mutate(pathway_clean = pathway %>%
           stringr::str_remove("^GOBP_") %>%
           stringr::str_replace_all("_", " ") %>%
           stringr::str_to_lower() %>%
           stringr::str_to_title()) %>%
  # Only name pathways with padj < 0.01 and |NES| > 2
  mutate(pathway_plot = ifelse(abs(NES) > 2 & padj < 0.01, pathway_clean, ""))

go_bp_plt_total <- ggplot(go_bp.total, aes(x = NES, y = SignedLogPadj, color = Category, label = pathway_plot)) +
  geom_point(size = 2) +
  geom_hline(yintercept = -log10(0.05), linetype='dotted', col = 'grey20') +
  geom_hline(yintercept = log10(0.05), linetype='dotted', col = 'grey20') +
  geom_label_repel(max.overlaps = 10) +
  theme_bw() +
  scale_color_manual(values = c("steelblue4", "grey70", "tomato3")) +
  xlab("Normalized Enrichment Score") +
  ylab("Signed -log10(adjusted p-value)") +
  ggtitle("Total RNA") +
  theme(plot.title = element_text(hjust = 0.5, face = "bold"))

go_bp_plt_labeled <- ggplot(go_bp.labeled, aes(x = NES, y = SignedLogPadj, color = Category, label = pathway_plot)) +
  geom_point(size = 2) +
  geom_hline(yintercept = -log10(0.05), linetype='dotted', col = 'grey20') +
  geom_hline(yintercept = log10(0.05), linetype='dotted', col = 'grey20') +
  geom_label_repel(max.overlaps = 10) +
  theme_bw() +
  scale_color_manual(values = c("steelblue4", "grey70", "tomato3")) +
  xlab("Normalized Enrichment Score") +
  ylab("Signed -log10(adjusted p-value)") +
  ggtitle("Labeled RNA") +
  theme(plot.title = element_text(hjust = 0.5, face = "bold"))

pdf(paste0(figs_dir, "MYC_vs_GFP_fgsea_go-bp_ranks_total.pdf"), height = 6, width = 6)
go_bp_plt_total
dev.off()

pdf(paste0(figs_dir, "MYC_vs_GFP_fgsea_go-bp_ranks_labeled.pdf"), height = 6, width = 6)
go_bp_plt_labeled
dev.off()

# Heatmap to compare labeled vs total
go_bp.nes.total <- go_bp.total %>%
  filter(padj < 0.01 & abs(NES) > 2) %>%
  select(pathway_clean, NES) %>%
  reshape2::melt()
go_bp.nes.total$variable <- "Total"

go_bp.nes.labeled <- go_bp.labeled %>%
  filter(padj < 0.01 & abs(NES) > 2) %>%
  select(pathway_clean, NES) %>%
  reshape2::melt()
go_bp.nes.labeled$variable <- "Labeled"

go_bp.compare <- rbind(go_bp.nes.total, go_bp.nes.labeled)

go_bp_compare_hmap <- ggplot(go_bp.compare, aes(x = variable, y = pathway_clean, fill = value)) + 
  geom_tile(color = "black") + 
  theme_bw() + 
  scale_fill_gradient2(low = "steelblue4", mid = "white", high = "tomato3", name = "NES") +
  xlab("RNA") +
  ylab("GO biological process")

pdf(paste0(figs_dir, "MYC_vs_GFP_fgsea_go-bp_comparison_hmap.pdf"), height = 9, width = 10)
go_bp_compare_hmap
dev.off()

go_bp_plt_total
go_bp_plt_labeled
go_bp_compare_hmap
```


```{r enrichment_go_mf}
go_mf.total <- fgsea(pathways = go_mf, stats = ranks.total, minSize  = 10, maxSize  = 500)
go_mf.labeled <- fgsea(pathways = go_mf, stats = ranks.labeled, minSize  = 10, maxSize  = 500)

go_mf.total$leadingEdge = sapply(go_mf.total$leadingEdge, function(x) paste0(x, collapse = ";"))
go_mf.labeled$leadingEdge = sapply(go_mf.labeled$leadingEdge, function(x) paste0(x, collapse = ";"))

write.table(go_mf.total, paste0(results_dir, "MYC_vs_GFP_gsea_go-mf_total.tsv"), col.names = TRUE, row.names = FALSE, quote = FALSE, sep = "\t")
write.table(go_mf.labeled, paste0(results_dir, "MYC_vs_GFP_gsea_go-mf_labeled.tsv"), col.names = TRUE, row.names = FALSE, quote = FALSE, sep = "\t")

# Rank plot: label pathways with padj < 0.01 and |NES| > 2
go_mf.total <- go_mf.total %>%
  # Create signed log(padj) column for rank plot
  mutate(SignedLogPadj = -log10(padj)) %>%
  mutate(SignedLogPadj = ifelse(NES > 0, SignedLogPadj, -SignedLogPadj)) %>%
  # Create category column for coloring
  mutate(Category = case_when(
    padj < 0.01 & NES > 0 ~ "UP",
    padj < 0.01 & NES < 0 ~ "DOWN",
    TRUE ~ "ns")) %>%
  # Clean pathway names
  mutate(pathway_clean = pathway %>%
           stringr::str_remove("^GOMF_") %>%
           stringr::str_replace_all("_", " ") %>%
           stringr::str_to_lower() %>%
           stringr::str_to_title())%>%
  # Only name pathways with padj < 0.01 and |NES| > 2
  mutate(pathway_plot = ifelse(abs(NES) > 2 & padj < 0.01, pathway_clean, ""))

go_mf.labeled <- go_mf.labeled %>%
  # Create signed log(padj) column for rank plot
  mutate(SignedLogPadj = -log10(padj)) %>%
  mutate(SignedLogPadj = ifelse(NES > 0, SignedLogPadj, -SignedLogPadj)) %>%
  # Create category column for coloring
  mutate(Category = case_when(
    padj < 0.01 & NES > 0 ~ "UP",
    padj < 0.01 & NES < 0 ~ "DOWN",
    TRUE ~ "ns")) %>%
  # Clean pathway names
  mutate(pathway_clean = pathway %>%
           stringr::str_remove("^GOMF_") %>%
           stringr::str_replace_all("_", " ") %>%
           stringr::str_to_lower() %>%
           stringr::str_to_title()) %>%
  # Only name pathways with padj < 0.01 and |NES| > 2
  mutate(pathway_plot = ifelse(abs(NES) > 2 & padj < 0.01, pathway_clean, ""))

go_mf_plt_total <- ggplot(go_mf.total, aes(x = NES, y = SignedLogPadj, color = Category, label = pathway_plot)) +
  geom_point(size = 2) +
  geom_hline(yintercept = -log10(0.05), linetype='dotted', col = 'grey20') +
  geom_hline(yintercept = log10(0.05), linetype='dotted', col = 'grey20') +
  geom_label_repel(max.overlaps = 10) +
  theme_bw() +
  scale_color_manual(values = c("steelblue4", "grey70", "tomato3")) +
  xlab("Normalized Enrichment Score") +
  ylab("Signed -log10(adjusted p-value)") +
  ggtitle("Total RNA") +
  theme(plot.title = element_text(hjust = 0.5, face = "bold"))

go_mf_plt_labeled <- ggplot(go_mf.labeled, aes(x = NES, y = SignedLogPadj, color = Category, label = pathway_plot)) +
  geom_point(size = 2) +
  geom_hline(yintercept = -log10(0.05), linetype='dotted', col = 'grey20') +
  geom_hline(yintercept = log10(0.05), linetype='dotted', col = 'grey20') +
  geom_label_repel(max.overlaps = 10) +
  theme_bw() +
  scale_color_manual(values = c("steelblue4", "grey70", "tomato3")) +
  xlab("Normalized Enrichment Score") +
  ylab("Signed -log10(adjusted p-value)") +
  ggtitle("Labeled RNA") +
  theme(plot.title = element_text(hjust = 0.5, face = "bold"))

pdf(paste0(figs_dir, "MYC_vs_GFP_fgsea_go-mf_ranks_total.pdf"), height = 6, width = 6)
go_mf_plt_total
dev.off()

pdf(paste0(figs_dir, "MYC_vs_GFP_fgsea_go-mf_ranks_labeled.pdf"), height = 6, width = 6)
go_mf_plt_labeled
dev.off()

# Heatmap to compare labeled vs total
go_mf.nes.total <- go_mf.total %>%
  filter(padj < 0.01 & abs(NES) > 2) %>%
  select(pathway_clean, NES) %>%
  reshape2::melt()
go_mf.nes.total$variable <- "Total"

go_mf.nes.labeled <- go_mf.labeled %>%
  filter(padj < 0.01 & abs(NES) > 2) %>%
  select(pathway_clean, NES) %>%
  reshape2::melt()
go_mf.nes.labeled$variable <- "Labeled"

go_mf.compare <- rbind(go_mf.nes.total, go_mf.nes.labeled)

go_mf_compare_hmap <- ggplot(go_mf.compare, aes(x = variable, y = pathway_clean, fill = value)) + 
  geom_tile(color = "black") + 
  theme_bw() + 
  scale_fill_gradient2(low = "steelblue4", mid = "white", high = "tomato3", name = "NES") +
  xlab("RNA") +
  ylab("GO molecular function")

pdf(paste0(figs_dir, "MYC_vs_GFP_fgsea_go-mf_comparison_hmap.pdf"), height = 8, width = 10)
go_mf_compare_hmap
dev.off()

go_mf_plt_total
go_mf_plt_labeled
go_mf_compare_hmap
```


```{r enrichment_go_cc}
go_cc.total <- fgsea(pathways = go_cc, stats = ranks.total, minSize  = 10, maxSize  = 500)
go_cc.labeled <- fgsea(pathways = go_cc, stats = ranks.labeled, minSize  = 10, maxSize  = 500)

go_cc.total$leadingEdge = sapply(go_cc.total$leadingEdge, function(x) paste0(x, collapse = ";"))
go_cc.labeled$leadingEdge = sapply(go_cc.labeled$leadingEdge, function(x) paste0(x, collapse = ";"))

write.table(go_cc.total, paste0(results_dir, "MYC_vs_GFP_gsea_go-cc_total.tsv"), col.names = TRUE, row.names = FALSE, quote = FALSE, sep = "\t")
write.table(go_cc.labeled, paste0(results_dir, "MYC_vs_GFP_gsea_go-cc_labeled.tsv"), col.names = TRUE, row.names = FALSE, quote = FALSE, sep = "\t")

# Rank plot: label pathways with padj < 0.01 and |NES| > 2
go_cc.total <- go_cc.total %>%
  # Create signed log(padj) column for rank plot
  mutate(SignedLogPadj = -log10(padj)) %>%
  mutate(SignedLogPadj = ifelse(NES > 0, SignedLogPadj, -SignedLogPadj)) %>%
  # Create category column for coloring
  mutate(Category = case_when(
    padj < 0.01 & NES > 0 ~ "UP",
    padj < 0.01 & NES < 0 ~ "DOWN",
    TRUE ~ "ns")) %>%
  # Clean pathway names
  mutate(pathway_clean = pathway %>%
           stringr::str_remove("^GOCC_") %>%
           stringr::str_replace_all("_", " ") %>%
           stringr::str_to_lower() %>%
           stringr::str_to_title())%>%
  # Only name pathways with padj < 0.01 and |NES| > 2
  mutate(pathway_plot = ifelse(abs(NES) > 2 & padj < 0.01, pathway_clean, ""))

go_cc.labeled <- go_cc.labeled %>%
  # Create signed log(padj) column for rank plot
  mutate(SignedLogPadj = -log10(padj)) %>%
  mutate(SignedLogPadj = ifelse(NES > 0, SignedLogPadj, -SignedLogPadj)) %>%
  # Create category column for coloring
  mutate(Category = case_when(
    padj < 0.01 & NES > 0 ~ "UP",
    padj < 0.01 & NES < 0 ~ "DOWN",
    TRUE ~ "ns")) %>%
  # Clean pathway names
  mutate(pathway_clean = pathway %>%
           stringr::str_remove("^GOCC_") %>%
           stringr::str_replace_all("_", " ") %>%
           stringr::str_to_lower() %>%
           stringr::str_to_title()) %>%
  # Only name pathways with padj < 0.01 and |NES| > 2
  mutate(pathway_plot = ifelse(abs(NES) > 2 & padj < 0.01, pathway_clean, ""))

go_cc_plt_total <- ggplot(go_cc.total, aes(x = NES, y = SignedLogPadj, color = Category, label = pathway_plot)) +
  geom_point(size = 2) +
  geom_hline(yintercept = -log10(0.05), linetype='dotted', col = 'grey20') +
  geom_hline(yintercept = log10(0.05), linetype='dotted', col = 'grey20') +
  geom_label_repel(max.overlaps = 10) +
  theme_bw() +
  scale_color_manual(values = c("steelblue4", "grey70", "tomato3")) +
  xlab("Normalized Enrichment Score") +
  ylab("Signed -log10(adjusted p-value)") +
  ggtitle("Total RNA") +
  theme(plot.title = element_text(hjust = 0.5, face = "bold"))

go_cc_plt_labeled <- ggplot(go_cc.labeled, aes(x = NES, y = SignedLogPadj, color = Category, label = pathway_plot)) +
  geom_point(size = 2) +
  geom_hline(yintercept = -log10(0.05), linetype='dotted', col = 'grey20') +
  geom_hline(yintercept = log10(0.05), linetype='dotted', col = 'grey20') +
  geom_label_repel(max.overlaps = 10) +
  theme_bw() +
  scale_color_manual(values = c("steelblue4", "grey70", "tomato3")) +
  xlab("Normalized Enrichment Score") +
  ylab("Signed -log10(adjusted p-value)") +
  ggtitle("Labeled RNA") +
  theme(plot.title = element_text(hjust = 0.5, face = "bold"))

pdf(paste0(figs_dir, "MYC_vs_GFP_fgsea_go-cc_ranks_total.pdf"), height = 6, width = 6)
go_cc_plt_total
dev.off()

pdf(paste0(figs_dir, "MYC_vs_GFP_fgsea_go-cc_ranks_labeled.pdf"), height = 6, width = 6)
go_cc_plt_labeled
dev.off()

# Heatmap to compare labeled vs total
go_cc.nes.total <- go_cc.total %>%
  filter(padj < 0.01 & abs(NES) > 2) %>%
  select(pathway_clean, NES) %>%
  reshape2::melt()
go_cc.nes.total$variable <- "Total"

go_cc.nes.labeled <- go_cc.labeled %>%
  filter(padj < 0.01 & abs(NES) > 2) %>%
  select(pathway_clean, NES) %>%
  reshape2::melt()
go_cc.nes.labeled$variable <- "Labeled"

go_cc.compare <- rbind(go_cc.nes.total, go_cc.nes.labeled)

go_cc_compare_hmap <- ggplot(go_cc.compare, aes(x = variable, y = pathway_clean, fill = value)) + 
  geom_tile(color = "black") + 
  theme_bw() + 
  scale_fill_gradient2(low = "steelblue4", mid = "white", high = "tomato3", name = "NES") +
  xlab("RNA") +
  ylab("GO cellular compartment")

pdf(paste0(figs_dir, "MYC_vs_GFP_fgsea_go-cc_comparison_hmap.pdf"), height = 6, width = 6)
go_cc_compare_hmap
dev.off()

go_cc_plt_total
go_cc_plt_labeled
go_cc_compare_hmap
```
```{r heatmap_selected_genes, echo=FALSE}
genes <- c("Myh6", "Ryr2", "Myc", "Mycn", "Yap1", "Aurka", "Ddx21", "Rrp9")

example.total <- scale(t(assay(vsd.total)))
example.total <- example.total[, genes]

example_total_hmap <- Heatmap(t(example.total), 
                        name = "z-score", 
                        show_column_names = FALSE, 
                        top_annotation = column_ha, 
                        show_row_names = TRUE, 
                        column_title = "Total RNA")

pdf(paste0(figs_dir, "MYC_vs_GFP_example-genes_hmap_total.pdf"), height = 3, width = 4)
example_total_hmap
dev.off()

example.labeled <- scale(t(assay(vsd.labeled)))
example.labeled <- example.labeled[, genes]

example_labeled_hmap <- Heatmap(t(example.labeled), 
                        name = "z-score", 
                        show_column_names = FALSE, 
                        top_annotation = column_ha, 
                        show_row_names = TRUE,
                        column_title = "Labeled RNA")

pdf(paste0(figs_dir, "MYC_vs_GFP_example-genes_hmap_labeled.pdf"), height = 3, width = 4)
example_labeled_hmap
dev.off()

example_total_hmap
example_labeled_hmap
```
```{r closing}
sessionInfo()
```
